---
title: "GC_RepData_PROJ1"
author: "Guillermo Cuenca"
date: "23 de junio de 2016"
output: GCPA1.html
---

## Introduction
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals throughout the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.
This document presents the results from Project Assignment 1 in the Coursera course Reproducible Research, written in a single R markdown document that can be processed by knitr and transformed into an HTML file.

## R preparations
In this document code will be represented to show how the results have been achieved. Set the default of echo to be true throughout the document:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load in packages

```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
```

## Working Directory

```{r}
setwd('C:/Users/gcuenca/Documents/01 DATA SCIENCE/05 REPRODUCIBLE RESEARCH/WEEK 2/PROJECT WEEK 2')
```

##The data

Show any code that is needed to:
1. Load the data (i.e. read.csv())
2. Process/transform the data (if necessary) into a format suitable for your analysis
Reading in Data
Data is loaded using the read.csv(), and stored in alldata.

```{r}
alldata <- read.csv("activity.csv", header = TRUE, sep = ',', colClasses = c("numeric", "character",   "integer"))
```

#Tidying the data

Change the date into dateformat using lubridate:
```{r}
alldata$date <- ymd(alldata$date)
```

#Check the data with str() and head():

```{r}
str(alldata)
head(alldata)
```

OK

## What is mean total number of steps taken per day?

For this part of the assignment the missing values can be ignored.
1. Calculate the total number of steps taken per day.
2. Make a histogram of the total number of steps taken each day.
3. Calculate and report the mean and median of the total number of steps taken per day.
Then
1. Calculate the total number of steps per day using dplyr and group by date, and store in steps:

```{r}
steps <- alldata %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print
```

2. Use ggplot for making the histogram:

```{r}
ggplot(steps, aes(x = steps)) +
  geom_histogram(fill = "firebrick", binwidth = 1000) +
  labs(title = "Histogram of Steps per day", x = "Steps per day", y = "Frequency")
```

3. Calculate the mean and median of the total number of steps taken per day and store in mean_steps & median_steps:

```{r}
mean_steps <- mean(steps$steps, na.rm = TRUE)
median_steps <- median(steps$steps, na.rm = TRUE)

## Mean steps are
mean_steps
##Median steps are
median_steps
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?.

Then

1. Calculate the average number of steps taken in each 5-minute interval per day using dplyr and group by interval, and stored in interval:

```{r}
interval <- alldata %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps))
```


Use ggplot for making the time series of the 5-minute interval and average steps taken:

```{r}
ggplot(interval, aes(x=interval, y=steps)) +
  geom_line(color = "firebrick")
```

2. Use which.max() to find out the maximum steps, on average, across all the days:

```{r}
interval[which.max(interval$steps),]
```

##Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Then

1. Summarize all the missing values:

```{r}
sum(is.na(alldata$steps))
```

2. Let’s take the approach to fill in a missing NA with the average number of steps in the same 5-min interval.

3. Create a new dataset as the original and use tapply for filling in the missing values with the average number of steps per 5-minute interval first stored alldata in data_full:

```{r}
data_full <- alldata
nas <- is.na(data_full$steps)
avg_interval <- tapply(data_full$steps, data_full$interval, mean, na.rm=TRUE, simplify=TRUE)
data_full$steps[nas] <- avg_interval[as.character(data_full$interval[nas])]
```

Check that there are no missing values:

```{r}
sum(is.na(data_full$steps))
```
No more missing values.

4. Calculate the number of steps taken in each 5-minute interval per day using dplyr and group by interval. Use ggplot for making the histogram:

```{r}
steps_full <- data_full %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print
ggplot(steps_full, aes(x = steps)) +
  geom_histogram(fill = "firebrick", binwidth = 1000) +
  labs(title = "Histogram of Steps per day, including missing values", x = "Steps per day", y = "Frequency")
```
Calculate the mean and median steps with the filled in values:

```{r}
mean_steps_full <- mean(steps_full$steps, na.rm = TRUE)
median_steps_full <- median(steps_full$steps, na.rm = TRUE)
mean_steps_full
median_steps_full
```

The impact of imputing missing data with the average number of steps in the same 5-min interval is the mean and the median values and are same.

## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() will come handy. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

Then 

1. Use dplyr and mutate to create a new column, weektype, and apply whether the day is weekend or weekday:

```{r}
data_full <- mutate(data_full, weektype = ifelse(weekdays(data_full$date) == "Saturday" | weekdays(data_full$date) == "Sunday", "weekend", "weekday"))
data_full$weektype <- as.factor(data_full$weektype)
head(data_full)
```

2. Calculate the average steps in the 5-minute interval and use ggplot for making the time series of the 5-minute interval for weekday and weekend, and compare the average steps:

```{r}
interval_full <- data_full %>%
  group_by(interval, weektype) %>%
  summarise(steps = mean(steps))
s <- ggplot(interval_full, aes(x=interval, y=steps, color = weektype)) +
  geom_line() +
  facet_wrap(~weektype, ncol = 1, nrow=2)
print(s)
```

It seems that the test object is more active earlier in the day during weekdays compared to weekends, but more active throughout the weekends compared with weekdays.

